Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing

### 2020-08-17 Dissonance_Santiago_v2/Dissonance_Santiago_Title_Page_v2.tex
v2 Revisions with BML

  \journal{Journal of Transport \& Health} % Sets Journal name

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance
from the Perspective of Affective Values}
    \author[Some University]{Author 1\corref{Corresponding Author}}
   \ead{a1@university.com} 
    \author[Some School]{Author 2}
   \ead{a2@school.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Some School]{Department, Street, City, State, Zip}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Transportation planning for decades has focused on providing mobility
for the private automobile. This is a model of development that was
initially introduced in North America as a solution to problems created
by rapid urbanization, and that was eventually copied elsewhere
(Angotti, 1996; Brown et al., 2009). Despite the initial promise of
automotive technology, it is now evident that mobility centered on the
private automobile has given rise to a litany of maladies that are in
urgent need of correction. This includes environmental concerns (i.e.,
climate change; Chapman, 2007) as well as numerous other social
(Boschmann and Kwan, 2008; Lucas, 2019, 2012), health (Khreis et al.,
2016; Milne, 2012), and equity issues (Bocarejo and Oviedo, 2012;
Martens et al., 2012; Pereira et al., 2017).

As the impacts of our societal dependence on the private automobile have
become increasingly evident, the transportation agenda has aimed to
shift focus to the reduction of car use and towards the creation of
mobility polycultures that offer a broader menu of transportation
alternatives other than primarily (or even just) the private automobile
(Lavery et al., 2013; Miller, 2011). In order to successfully achieve
this goal, it is essential not only to provide the services and
facilities that support public transportation and active travel, but
also to attract new users to these modes of transportation (Ettema et
al., 2011). Within this context, it has been argued that sustainable
transportation policies require all participants in the transportation
system to challenge what Gossling and Cohen (2014) termed
\emph{transportation taboos}: deep-seated ideas concerning the
contribution to emissions by individuals, the inequality of market-based
approaches, and the social and psychological functions of
transportation. With respect to the latter, subjective wellbeing (SWB)
is increasingly recognized as a policy goal in and of itself (Diener et
al., 2015; Dolan and Metcalfe, 2012), and also as an intermediate
objective to achieve ulterior goals. Accordingly, there is a growing
consensus in the field of transportation about the need to move beyond a
purely utilitarian focus to understand the affective and wellbeing
values of the transportation experience (e.g., Anable and Gatersleben,
2005; De Vos et al., 2013; Domarchi et al., 2008; Gatersleben and
Uzzell, 2007; Steg, 2005). Consideration of factors related to
subjective wellbeing, for example, are useful for policy-makers to
enhance the experience of travel, to increase satisfaction with travel,
and to meet the preferences of travelers (Chatterjee et al., 2019).

A definition of subjective wellbeing is as ``{[}g{]}ood mental states,
including all of the various evaluations, positive and negative, that
people make of their lives, and the affective reactions of people to
their experiences'' (OECD, 2013, p. 10). As memorably put by Steg
(2005): is the use of a car a must, or is it a lust? A close alignment,
or consonance, between affective reactions and the mode of
transportation used to travel can result in greater subjective wellbeing
and increase the probability of choosing that mode (i.e., by satisfying
the \emph{lust}); on the other hand, dissonance, namely the lack of
concordance between the mode used and the affective reaction to that
mode, would be detrimental to subjective wellbeing, especially to those
for whom the use of the mode is a \emph{must} (De Vos, 2019; Mokhtarian
and Pendyala, 2018). With these considerations in mind, the present
research aims to investigate \emph{who} experiences affective dissonance
with respect to their primary mode of travel. We focus our investigation
on reactions from the perspective of feelings of freedom, enjoyment,
happiness, poverty, luxury, and status - concepts termed \emph{affective
values} hereafter. Furthermore, it is possible, even if use of a
particular mode is a \emph{must}, that travelers may still \emph{lust}
for something else - in other words, the grass may look greener from the
window of a car. For this reason, we also aim to investigate which mode
or modes are most commonly identified as ideal from the perspective of
the aforementioned affective values.

Our research is based on data collected from a sample of travelers in
the city of Santiago in Chile. The analysis sheds light on the affective
reactions of people towards their primary mode of travel, and their
perceptions towards `ideal modes'. This reveals important differences in
situations when the ideal mode of a traveler is not the mode they
actually use - or in other words, the affective dissonance between
actual and ideal modes of travel. More concretely, the results indicate
that users of public transportation experience the most dissonance,
followed by automobile users, and finally active travelers, who tend to
experience the least affective dissonance. For those travelers who
experience dissonance, active travel is the mode most commonly
associated with freedom, enjoyment, and happiness, whereas public
transportation is most commonly associated with poverty, and the
automobile is most commonly associated with luxury and status. We also
find that there are significant variations in dissonance by age,
education, income, and typical commute time. The reactions of travelers
towards various transport modes are critical factors that policy-makers
need to consider to promote and increase the use of public transport and
active modes (Bornioli et al., 2019; De Vos et al., 2019; De Vos and
Witlox, 2017; Garling et al., 2019; Redman et al., 2013).

After these introductory remarks, the paper follows up with a review of
the literature on the topics of subjective wellbeing and affective
values. Next, we discuss the case study, as well as the methodology and
data used for this research. The analysis and results of the study are
presented afterwards, before concluding with some discussion and
directions for future research and policy.

Please note that this document was prepared using R Markdown and is an
example of reproducible research (Brunsdon and Comber, 2020). The R
Markdown file, along with the data file needed to reproduce the
analysis, are available for
download\footnote{\url{https://drive.google.com/open?id=189ZvfVvRis5xZA9IlviKCLW2bwbPJAgg}}.

\hypertarget{background}{%
\section{Background}\label{background}}

A consensus has emerged in recent years regarding the need to complement
the traditional utilitarian perspective of transportation by
investigating mobility and transport from the lens of their affective
functions. The affective value of transportation is important due to its
potential to improve or detract from SWB.

One of the primary ways to explore issues of subjective wellbeing in
transportation has been the satisfaction that travelers feel towards
their everyday mobility experience (e.g., Cecilia Jakobsson Bergstad et
al., 2011). As a consequence, there is a wealth of research on
satisfaction with the use of different modes of transportation. For
example, numerous studies report that car users often have a higher
level of satisfaction compared to other transport modes (C. J. Bergstad
et al., 2011; Eriksson et al., 2013; Redmond and Mokhtarian, 2001),
although others report highest commute satisfaction for bicycle and
train commuters (e.g., Whalen et al., 2013; Handy and Thigpen, 2019). In
a similar way, there are multiple reports that active travel also tends
to yield higher levels of satisfaction (Gatersleben and Uzzell, 2007;
Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith, 2017; St-Louis et
al., 2014; Whalen et al., 2013), In contrast, public transport users
often assess their experience more negatively (Abenoza et al., 2017; De
Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy and Thigpen, 2019;
Paez and Whalen, 2010). Multi-modal trips also influence satisfaction
levels; for instance, individuals tend to report higher levels of
satisfaction with the mode they use -- perhaps as a form of \emph{post
hoc} validation of their choices or circumstance (Susilo and Cats,
2014).

While travel satisfaction has often been used in the context of daily
trips -- typically linked to cost-benefit and utilitarian measurements
--, the evaluation of subjective wellbeing (SWB) over time has risen as
an alternative measure. Ettema et al.~(2010, p. 725) define SWB as the
degree to which an individual positively evaluates the overall quality
of their lives, where the general life satisfaction encompasses a more
extended temporal period -- which implies greater temporal stability.
This concept has prompted a growing literature that complements and
applies SWB in a broader range of satisfaction scales and situations.
The definition of other factors such as travel choice mode, attitudes,
and external elements of the built environment have been studied for a
broader understanding of the changes produced in the SWB (e.g., Handy
and Thigpen, 2019). As these factors do not necessarily apply to
longer-term general life satisfaction, other studies have aimed to
determine both the direct and indirect effects on the perception of
users (see, e.g., Ye and Titheridge, 2017). Other concepts have also
emerged, including the Satisfaction with Travel Scale (Ettema et al.,
2011), as well as different scales based on people's perceptions of
travel. De Vos et al.~(2015), for instance, explore in detail the
underlying dimensions of the affective domain of STS on which SWB is
based (for more on STS also see Friman et al., 2013).

The literature on SWB has demonstrated a relationship between people's
perceptions and satisfaction with their daily travel (e.g., Smith, 2017;
Mokhtarian and Pendyala, 2018; St-Louis et al., 2014). Scholars have
shown that accessibility is among the most studied factors that
influences subjective wellbeing (Delbosc, 2012), and activities have a
direct impact on travel satisfaction (Cecilia Jakobsson Bergstad et al.,
2011). Delbosc (2012, p. 28), for instance, has summarized the most
significant influences on psychological wellbeing: poverty and
employment, meaningful relationships and health. However, understanding
the components that affect these perceptions implies the differentiation
between affective (also named as symbolic-affective) and instrumental
values (C. J. Bergstad et al., 2011). Steg et al.~(2011) have compared
symbolic-affective and instrumental-reasoned motives for car-use, and
similar findings are reported in other studies (e.g., Gatersleben and
Uzzell, 2007; Lois and Lopez-Saez, 2009). Previous research also
demonstrates how socio-demographic factors affect the levels of SWB,
includin income (Clark and Oswald, 1996; Ferrer-i-Carbonell, 2005),
education and unemployment (Argyle et al., 1999), age (Diener and
Eunkook Suh, 1997), and gender (Tesch-Römer et al., 2008). However,
further research beyond the determinants of trip satisfaction is needed
to understand how these socio-demographic variables connect with the
affective reactions to various modes of travel (St-Louis et al., 2014).

The research needs outlined above are well-recognized in the developed
world, but there is still a dearth of research in the context of the
Global South. Historical inequality in many developing countries has led
to strong symbolic attachments to the automobile, in addition to
negative connotations for public transport and active travel (Zorrilla
et al., 2019). To address this gap in knowledge a body of literature has
emerged to investigate the affective aspects of travel behavior in a
number of developing countries. A cross-country study in Asia revealed
that the affective factors of public transportation and car use are
important, and in particular the social orderliness of transit was
suggested as a way to make this mode more attractive to users (Van et
al., 2014). In terms of active travel, a study in China found that
attitudes that embrace new styles and technologies despite their cost
are associated with the intention to continue using shared bicycles
(Shao and Liang, 2019). The importance of affective factors for policy
and planning is further highlighted by research in Colombia that shows
the pride users feel when using a bicycle shared system. This affective
reaction is above and beyond other positive feelings, such as a sense of
belongingness to a civic culture, and the enjoyment and pleasure of
active travel itself (Bejarano et al., 2017).

\hypertarget{case-study-and-data}{%
\section{Case Study and Data}\label{case-study-and-data}}

\hypertarget{context}{%
\subsection{Context}\label{context}}

The case study for the present research is Santiago. Santiago is the
capital of Chile, a country with one of the highest levels of
socio-economic inequality in the world. A tangible manifestation of the
inequalities experienced by many in Chile is the large disparity in the
relative cost of transportation, travel time, and distance traveled by
different socio-economic segments of the population. To further
complicate matters, strong spatial segregation also conditions car
ownership and the use of public transportation -- the higher the income,
the higher the use of the automobile; conversely, the lower the income,
the higher the reliance on public transportation. Although the
transportation experience is but one of many dimensions of inequality,
the experiences in this sector have triggered exceptional discomfort and
dissatisfaction among the public. Recent social unrest, triggered by a
seemingly minor hike in the fare of public transportation, has brought
many of these concerns to the forefront of the public conscience in
dramatic fashion (e.g., Davies, 2019; {\textbf{???}}; {\textbf{???}}).

Previous research has helped to contextualize everyday mobility in
Santiago relative to other Latin American cities (e.g., Avellaneda and
Lazo, 2011; Rodríguez Vignoli, 2008), but much remains to be explored.
Measuring instruments and new methods have led to more accurate and
precise understandings of the social issues that arise as consequence of
transport infrastructure and housing provision (Cox and Hurtubia, 2016),
and the minimum provision of basic services (Tiznado-Aitken et al.,
2016). However, the focus on accessibility as a measure of inequalities
remains predominant ({\textbf{???}}; Martínez et al., 2018; Niehaus et
al., 2016; Rojas et al., 2016).

\hypertarget{data}{%
\subsection{Data}\label{data}}

The study is based on a survey conducted in the city of Santiago during
the months of November and December of 2016, at the end of the Spring
and beginning of Summer seasons. The survey collected information on a
wide range of transportation and related issues, and the data collection
protocol considered a quota-sampling method for gathering the
information, considering the socio-demographic information from
Pre-Census of 2012. The survey was carried out face-to-face in centers
of activity with dense provision of offices, services, and educational
centers. An equal representation of both genders and a representation of
the proportion of inhabitants per area were chosen as relevant
characteristics of the sample. In total, there were \(n=451\) valid
reponses, although not every response was complete and some questions
have missing values.

The study considers data from 3 out of 8 sections of a longer survey.
The relevant sections of the survey concern the individual
characteristics of respondents, their feelings and affective responses
related to their commute, and aspects describing their regular commute
trips. In terms of individual characteristics and their commute,
participants were asked socio-demographic information, including age
(coded in three classes: younger than 35, older than 35 but younger than
55, and 55 and olde); level of education (three classes: Kindergarden to
grade 12, or K-12, technical diploma or university graduate, and
graduate degree), income (three classes, by tertile), and the typical
duration of the respondent's regular commute (four classes, by duration
in minutes). The descriptive statistics of the sample appear in Figure
\ref{fig:descriptive-statistics}. The sample trends to younger (56\% is
younger than 34 years old) and well-educated respondents (68\% of
respondents having technical or professional education), with an almost
uniform distribution of income levels. The trend in typical commute time
is towards longer commutes -- e.g.~55\% of the trips are longer than 40
minutes, and from those 56\% corresponds to trips over the 60 minutes
long.

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/plot-descriptive-statistics-1.pdf}
\caption{\label{fig:descriptive-statistics}Descriptive statistics of the
sample}
\end{figure}

In addition, respondents were asked about their primary mode of travel
for their regular commute. The modes available were Car; Taxi;
\emph{Colectivo} (a form of shared ride, intermediate in flexibility and
capacity between taxi and bus); Motorcycle; Metro; Bus; Bicycle; and
Walking. As seen in the top panel of Figure
\ref{fig:primary-mode-travel}, the three most common modes of travel are
Metro, Bus, and Car, followed by Walking and Bicycle. For the analysis,
we aggregate these modes into the following categories (bottom panel of
Figure \ref{fig:primary-mode-travel}): (26\% of total modes -- which
corresponds to 91\% of the total of private motorized modes), Active
(9\% - Walking + Bicycle), Public (60\% - Metro + Bus), and Other
motorized modes (6\% - Taxi + Colectivo + Motorcycle).

\begin{figure}
\centering
\includegraphics{Dissonance_Santiago_v2_files/figure-latex/figure-primary-mode-travel-1.pdf}
\caption{\label{fig:primary-mode-travel}Frequency of primary mode used
for regular commute; top panel: all modes, bottom panel: aggregated
modes}
\end{figure}

Of particular interest for the present study is the following question
in Part 3 of the survey:

\begin{quote}
\begin{quote}
Q: Please indicate the mode(s) of transport that you relate to the
following feelings and concepts
\end{quote}
\end{quote}

The question was asked for each of the following affective values:
Freedom; Enjoyment; Happiness; Poverty; Luxury; and Status. The
respondents were not constrained to select only one alternative, but
could indicate by means of checkboxes any and all modes that they felt
aligned with each affective value. This allows us to analyze the mode
dissonance. Dissonance is a concept introduced into the transportation
literature by Schwanen and Mokhtarian (2004) based on earlier work by
Feldman (1990). Originally used to investigate residential dissonance
(Schwanen and Mokhtarian, 2004), the concept has since been extended in
the travel behavior literature to encompass the mismatch between the
choices individuals make, and the alternatives that would enable users
to improve their experience. This includes, for example, travel mode
dissonance (De Vos, 2018).

Based on the primary mode of travel and the questions about affective
values, we derived a series of travel mode dissonance variables
according to the following rule: \[
D_{ij} = 
\begin{cases}
0 & if \text{ primary mode of traveler }i == \text{mode associated with value } j\\
1 & \text{ otherwise}
\end{cases}
\] Therefore, if respondent \(i\) travels primarily by Car, but
indicated any other mode(s) in relation to Freedom, the respondent
experienced dissonance: \[
D_{i,\text{Freedom}} = 1
\]

In these calculations we account for all modes identified by respondents
in relation to the affective values. Therefore, to avoid double counting
the respondents in our frequency tabulations, we also calculated a
sample weight as the inverse of the number of modes selected in response
to each affective value. For instance, for respondents who selected two
modes in relation to affective value \(j\), each of these two modes
receive a weight of \(w_{ij}=1/2\); if a respondent selected three
modes, then their weights are \(w_{ij}=1/3\); and so on. In this way we
do not treat unfairly those respondents who selected only one mode, and
the sum of all weighted responses is equal to the size of the sample
\(n\).

\hypertarget{results}{%
\section{Results}\label{results}}

In what follows, the results refer to two related but distinct
questions. The first part of the analysis seeks to understand \emph{who}
experiences dissonance, whereas the second part, building off that, aims
to explore \emph{which modes} are more commonly identified as embodying
affective values by those travelers who experience dissonance.

\hypertarget{who-experiences-dissonance}{%
\subsection{Who experiences
dissonance?}\label{who-experiences-dissonance}}

We begin by profiling the travelers who experience dissonance. The null
hypothesis is that there are no systematic differences in terms of who
tends to experience affective dissonance with respect to their primary
mode of travel. To investigate this hypothesis we create contingency
tables to tabulate the frequency of dissonance with respect to each
affective value, stratified by the individual attributes of respondents.
Table \ref{tab:cross-tabulation-results} presents the frequency (in
percentage) of dissonance. This is the percentage of respondents out of
the total in their stratum who selected, for each affective value, a
mode or mode(s) of transportation that do not correspond to their
primary mode of travel.

As seen in the table, there are five individual attributes that we
analyze. Three of these are socio-economic and demographic, namely age,
level of education, and income. The other two are transportation
related, namely primary mode of travel and commute time. The frequency
tables are tested in every case using \(\chi^2\) tests of independence
(\(p\)-values are reported in the table; lower \(p\)-values mean that
the null hypothesis of independence can be rejected with greater
confidence). It is interesting to note that the only category for which
all affective values are significant at better than 5\% level of
confidence is mode of travel.

\hypertarget{age}{%
\subsubsection{Age}\label{age}}

With respect to age, previous studies have reported that older adults
tend to be more satisfied with their travel experience than younger
people (Cao and Ettema, 2014; De Vos et al., 2016; Ye and Titheridge,
2017). In the case, we find that there there are significant differences
in dissonance by age with respect to five affective values, namely
Freedom, Enjoyment, Happiness, Luxury, and Status. We observe that
levels of dissonance tend to be high in general, and in no case less
than 60\%. For instance, almost 90\% of travelers younger than 35
experience travel mode dissonance with respect to Freedom, and more than
94\% experience dissonance with respect to Enjoyment. In general,
younger travelers tend to experience dissonance more frequently, with
dissonance being less frequent for older travelers. The exception to
this trend is Luxury, an affective value for which older travelers (age
\textgreater54) experience dissonance more frequently than mid-aged
travelers (ages 35-54).

\hypertarget{education}{%
\subsubsection{Education}\label{education}}

The results do not allow us to reject the hypothesis of equal levels of
dissonance by levels of education for the values of Freedom, Enjoyment,
Happiness, and Poverty. In contrast we can reject the null hypothesis in
the case of two affective values, namely Luxury and Status. In the case
of Luxury, dissonance is more frequent among people who have only K-12
education, and less frequently, albeit still high, for people with
technical/university level education and post-graduate education.
Furthermore, highly educated people (with postgraduate degrees)
experience dissonance with respect to Status more frequently than with
respect to Luxury.

\hypertarget{income}{%
\subsubsection{Income}\label{income}}

The next individual attribute that we examine is income, and in this
case we detect significant differences in dissonance for three affective
values: Poverty, Luxury, and Status. Since Poverty is a negative affect,
we see here that lower income individuals tend to associate this feeling
to their primary mode of commuting with higher frequency (almost 20\% of
the time) than other income groups. Whereas approximately 16\% of
mid-income people are dissatisfied with their primary mode of travel
when it comes to evoking feelings of Poverty, less than 7\% of high
income individuals are. Dissonance with respect to Luxury and Status
also tends to be more common among lower income individuals, and
declines substantially for mid- and high income respondents. Notice as
well that the frequency of dissonance is higher in terms of Luxury than
Status for mid- and high income people.

\hypertarget{primary-mode-of-travel}{%
\subsubsection{Primary mode of travel}\label{primary-mode-of-travel}}

The variable that shows the largest differences in the frequency of
dissonance is the primary mode of travel. It can be seen in Table
\ref{tab:cross-tabulation-results} that the differences are significant
for all six affective values. Compared to users of other modes of
transportation, dissonance is particularly acute for users of public
transportation when it comes to the values of Freedom, Enjoyment, and
Happiness: almost 100\% of users of public transportation have
identified other mode or modes as better representing those affective
values. Dissonance on these values is the least for active travelers:
less than 50\% of respondents who are active travelers associate Freedom
to a different mode, and only around 60\% identified a different mode
when responding to the values of Enjoyment and Happiness, compared to
approximately 79\% and 71\% of those whose primary mode is car. The
picture changes when the values of Poverty, Luxury, and Status are
considered. In this case, dissonance is less frequent for people who
travel by car: less than 3\% of car users associate the car with
feelings of Poverty, only 41\% associate Luxury with a mode other than
car, and only about 31\% relate Status to a different mode. Dissonance
is more frequent in these values for active travelers, and users of
public transportation and other modes, in no case being lower than 75\%.
This figure is virtually as high as 100\% for users of public
transportation, a group of travelers who consistently associate Luxury
and Status with modes \emph{other} than public transportation.

\hypertarget{typical-commute-time}{%
\subsubsection{Typical commute time}\label{typical-commute-time}}

Turning now to typical commute time, four affective values show
significant differences at better than 10\% confidence: Freedom,
Happiness, Luxury, and Status. Perhaps not surprisingly, dissonance is
more frequent among people whose typical commutes are longer. This is in
line with previous findings: both St-Louis et al.~(2014) and Smith
(2017) report that commute satisfaction tends to decline with longer
commutes, whereas Handy and Thigpen (2019) found that commute distance
was a negative covariate of commute satisfaction.

\begin{landscape}\begin{table}

\caption{\label{tab:table-cross-tabulation-results-without-instrumental}\label{tab:cross-tabulation-results}Percentage of respondents who report mode dissonance with respect to various affective values}
\centering
\begin{tabular}[t]{lcccccccccccc}
\toprule
Variable & Freedom & $\chi^2$ p-val & Enjoyment & $\chi^2$ p-val & Happiness & $\chi^2$ p-val & Poverty & $\chi^2$ p-val & Luxury & $\chi^2$ p-val & Status & $\chi^2$ p-val\\
\midrule
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Age}}\\
\hspace{1em}< 35 & 89.88 &  & 94.33 &  & 93.52 &  & 87.63 &  & 89.43 &  & 87.17 & \\

\hspace{1em}35 - 54 & 74.22 &  & 81.45 &  & 82.26 &  & 87.38 &  & 70.18 &  & 68.70 & \\

> 54 & 74.00 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 72.92 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 68.75 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 76.19 & \multirow{-3}{*}{\centering\arraybackslash 0.4095} & 78.57 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$} & 66.67 & \multirow{-3}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Education}}\\
\hspace{1em}K-12 & 85.98 &  & 92.16 &  & 89.22 &  & 79.27 &  & 95.10 &  & 92.39 & \\

\hspace{1em}Tech/Univ & 83.00 &  & 86.94 &  & 86.94 &  & 88.51 &  & 78.99 &  & 75.00 & \\

Graduate & 78.57 & \multirow{-3}{*}{\centering\arraybackslash 0.9063} & 85.19 & \multirow{-3}{*}{\centering\arraybackslash 0.7005} & 84.62 & \multirow{-3}{*}{\centering\arraybackslash 0.9694} & 86.36 & \multirow{-3}{*}{\centering\arraybackslash 0.3608} & 76.00 & \multirow{-3}{*}{\centering\arraybackslash 0.0058} & 79.17 & \multirow{-3}{*}{\centering\arraybackslash 0.013}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Income}}\\
\hspace{1em}Low & 86.51 &  & 82.11 &  & 88.62 &  & 80.19 &  & 88.71 &  & 88.70 & \\

\hspace{1em}Middle & 84.52 &  & 89.80 &  & 88.00 &  & 83.76 &  & 85.82 &  & 80.00 & \\

High & 79.19 & \multirow{-3}{*}{\centering\arraybackslash 0.5755} & 90.97 & \multirow{-3}{*}{\centering\arraybackslash 0.2264} & 85.82 & \multirow{-3}{*}{\centering\arraybackslash 0.9698} & 93.69 & \multirow{-3}{*}{\centering\arraybackslash 0.0642} & 73.68 & \multirow{-3}{*}{\centering\arraybackslash 0.0204} & 70.31 & \multirow{-3}{*}{\centering\arraybackslash 0.0137}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Mode}}\\
\hspace{1em}Car & 58.93 &  & 78.90 &  & 70.91 &  & 97.96 &  & 41.00 &  & 30.69 & \\

\hspace{1em}Active & 46.34 &  & 60.98 &  & 57.89 &  & 75.76 &  & 89.47 &  & 81.82 & \\

\hspace{1em}Public & 99.23 &  & 96.76 &  & 98.80 &  & 81.35 &  & 100.00 &  & 99.57 & \\

Other & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $< 0.001$} & 86.96 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 91.30 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 93.33 & \multirow{-4}{*}{\centering\arraybackslash 0.0045} & 72.73 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 90.00 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$}\\
\cmidrule{1-13}
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Commute Time}}\\
\hspace{1em}< 20 min & 65.93 &  & 82.95 &  & 77.27 &  & 86.11 &  & 73.49 &  & 67.09 & \\

\hspace{1em}20 - 40 min & 85.86 &  & 87.50 &  & 89.58 &  & 90.41 &  & 83.87 &  & 81.18 & \\

\hspace{1em}40 - 60 min & 83.04 &  & 89.91 &  & 89.62 &  & 86.08 &  & 82.35 &  & 77.78 & \\

> 60 min & 95.93 & \multirow{-4}{*}{\centering\arraybackslash $<0.001$} & 92.17 & \multirow{-4}{*}{\centering\arraybackslash 0.6115} & 93.28 & \multirow{-4}{*}{\centering\arraybackslash 0.0394} & 83.02 & \multirow{-4}{*}{\centering\arraybackslash 0.9225} & 91.15 & \multirow{-4}{*}{\centering\arraybackslash 0.0925} & 90.09 & \multirow{-4}{*}{\centering\arraybackslash 0.0158}\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\hypertarget{which-modes-do-travelers-associate-with-affective-values}{%
\subsection{Which modes do travelers associate with affective
values?}\label{which-modes-do-travelers-associate-with-affective-values}}

The preceding analysis indicates that there is significant mode
dissonance along various dimensions and for various affective values.
This is for the most part in line with previous research, although by
examining different affective values individually instead of a summary
measure of wellbeing, we are able to differentiate better the affective
reactions of travelers. Less is known about the values that travelers
associate with modes \emph{other} than the one they use. For this
reason, after developing a profile of the travelers who experience mode
dissonance in the preceding section, we are now interested in the
responses of travelers with respect to the modes they ideally tend to
associate with various affective values. For this analysis we employ a
visualization technique known as \emph{faceting}. Bar charts are used to
plot the proportions of respondents who associate each mode of
transportation with an affective value. Faceting allows us to explore
these proportions in a multivariate way by slicing the data according to
additional attributes. The result is a visual representation of a
multi-way contingency table. Please note that due to the small numbers
of travellers who use modes categorized as ``Other'' we henceforth
exclude them from the analysis.

We begin our investigation of the modes more frequently associated with
different affective values by plotting the primary mode of travel and
the modes associated with the values (see Figure
\ref{fig:bar-plots-by-attribute}). We can think of the figure as a
matrix of plots, with each dimension of the matrix a \emph{facet}. In
this figure, the columns of the matrix of plots correspond to different
affective values, and the rows to different modes of primary mode of
transportation, i.e., the mode actually used. The height of the bars in
these plots is the proportion of respondents who selected mode
\(j = \text{Car, Active, Public}\) as embodying the corresponding
affective value.

We know from the preceding analysis (Table
\ref{tab:cross-tabulation-results}) that users of public transportation
tend to experience dissonance with significantly higher frequency than
car users and active travelers. But which mode(s) would align better to
their expectations with regards to the affective values under study? If
we inspect the bottom row of plots, we see that public transport users
most of the time associate Freedom, Enjoyment, and Happiness with active
travel. Users of public transportation are more or less equally split in
terms of associating Poverty with active travel and public
transportation, whereas they overwhelmingly associate Luxury and Status
with the private car.

Active travelers rarely experience dissonance with respect to Freedom,
Enjoyment, and Happiness. And although the experience of dissonance is
more frequent with respect to Luxury and Status, there is still a
sizeable plurality of active travellers who associate these two
affective values with active travel, \emph{not} the car (contrast these
to the responses of public transit users.)

Interestingly, dissonance among car users with respect to Freedom,
Enjoyment, and Happiness is relatively high. While car is still the most
frequently chosen mode associated with Freedom among car users, the
proportion of car users who identify Freedom with active travel is
almost as high. And when it comes to Enjoyment and Happiness, car users
experience significant dissonance and in fact a majority tend to
indicate active travel as representing these two affective values. In
contrast, car users display little dissonance with respect to feelings
of Luxury and Status.

It is interesting to note that feelings of ``Poverty'' are associated
with public transportation and active travel in almost exactly the same
proportions, irrespective of the primary mode of travel.

>>>>

  \journal{Journal of Transportation \& Health} % Sets Journal name

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header

@@

  \title{Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance
from the Perspective of Subjective Wellbeing}
    \author[PUC]{Beatriz Mella Lira\corref{Corresponding Author}}
   \ead{beatriz.lira.14@ucl.ac.uk} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[PUC]{BRT+ Centre of Excellence, Departamento Ingeniería de Transporte y
Logística, Escuela de Ingeniería, Pontificia Universidad Católica de
Chile}
    \address[McMaster University]{School of Earth, Environment, and Society, McMaster University, 1280
Main St W, Hamilton, Ontario, L8S 1S4 Canada}

@@
